CAMELOT Research Scholarship -- Final Report

### 2023-03-20 tex/report_fct/report.tex
Added FCT report

% Some of our commands for this guide.
%
\authorrunning{Abelt and Fonseca}

@@

%% Front Matter
%%
% Regular title as in the article class.
%
\title{LayeredTypes -- Combining dependent and independent type systems}

% Authors are joined by \and. Their affiliations are given by \inst, which indexes
% into the list defined using \institute
%
\author{
Lukas Abelt\inst{1,2}
\and
Alcides Fonseca\inst{1}
}

@@

% \titlerunning{} has to be set to either the main title or its shorter
% version for the running heads. When processed by
% EasyChair, this command is mandatory: a document without \titlerunning
% will be rejected by EasyChair
\titlerunning{Layered Types}

@@

%------------------------------------------------------------------------------
\section{Context}
\label{sec:context}
Most programmers and computer scientists will be familiar with simple type systems that ensure that the code they write is type safe in the context of the language they are writing it in. However, much more sophisticated type systems exist that can be used to ensure specific properties of a program, such as resource usage (Linear Types~\cite{linear}), value predicates (Liquid Types~\cite{DBLP:conf/pldi/RondonKJ08}), or correct following of a distributed protocol (Session Types~\cite{session}).

@@

 % Alcides is here

@@

%\begin{itemize}
%	\item From a type-safety perspective, one wants to make sure that e.g. \texttt{get(...)} is only called with a list and an integer index as arguments
%	\item Additionally, one wants to ensure that \texttt{get} is never accessed with an index that exceeds the lists length to avoid out-of-bounds access errors at runtime
%	\item When working with resources such as file descriptors one wants to ensure that these are in a proper (closed) state by the end of the program execution
%\end{itemize}

@@

\section{Proposed Approach}
\label{sec:proposed-approach}

We propose a second, more principled alternative: \LayeredTypes. In \LayeredTypes, developers can write programs, but can also define additional type systems as layers. Each layer defines the basic types, and how typechecking happens. Additionally, a layer may depend on another layer if it requires information (such as types or typing contexts) from another layer.

@@

We want to note an important distinction to Gradual Typing~\cite{gradual,gradual-objects}: Gradual Typing allows to only provide partial typing information in one system, but does not allow to choose different sets of types to verify. It allows to blend between having no types (0) and having types (1). Our proposal supports multiple type systems, with different combinations among them.

We evaluated our approach in a prototype language. Users can provide their own implementations for verification layers and define dependencies between them. In \Cref{lst:code_after} we see a set of annotations that can be added to the base code of \Cref{lst:code_before} to tackle the issues described. We define three separate layers \texttt{state}, \texttt{types} and \texttt{liquid}. Internally, the framework will build a dependency graph thus allowing to verify properties independently from one another where appropriate.

@@

>>>>

% Some of our commands for this guide.
%
\authorrunning{Abelt}

@@

%% Front Matter
%%
% Regular title as in the article class.
%
\title{CAMELOT Research Scholarship -- Final Report}

% Authors are joined by \and. Their affiliations are given by \inst, which indexes
% into the list defined using \institute
%
\author{
Lukas Abelt\inst{1,2}
}

@@

% \titlerunning{} has to be set to either the main title or its shorter
% version for the running heads. When processed by
% EasyChair, this command is mandatory: a document without \titlerunning
% will be rejected by EasyChair
\titlerunning{Project Report CAMELOT}

@@

%------------------------------------------------------------------------------
\section{Introduction}

This document is the final project report for the research scholarship granted in the scope of the CAMELOT project for the duration of 01.09.2022-30.11.2022. It was carried out by Lukas Abelt in the LASIGE Research Group within the Reliable Software Systems Research Line. The work was supercised by Alcides Fonseca, Assistant Professor at the Faculdade de Ciências da Universidade de Lisboa. The work was part of an Erasmus internship that ends in March 2023.

\section{Description of the Scholarship}
\subsection{Context}
\label{sec:context}
Most programmers and computer scientists will be familiar with simple type systems that ensure that the code they write is type safe in the context of the language they are writing it in. However, much more sophisticated type systems exist that can be used to ensure specific properties of a program, such as resource usage (Linear Types~\cite{linear}), value predicates (Liquid Types~\cite{DBLP:conf/pldi/RondonKJ08}), or correct following of a distributed protocol (Session Types~\cite{session}).

@@

@@

@@

\subsection{Goals}
\label{sec:proposed-approach}

The Goals of the Scholarship was to explore a new verification approach, called \LayeredTypes. The Goal was, that in \LayeredTypes, developers can write programs, but can also define additional type systems as layers. Each layer defines the basic types, and how typechecking happens. Additionally, a layer may depend on another layer if it requires information (such as types or typing contexts) from another layer.

@@

For this purpose, a prototype was to be developed consisting of a small and simple functional language accompanied by a prototype layered Verification tool that can realise the described layered verification approach.

@@

\subsection{Results obtained}
During the duration of the scholarship a prototype compiler and verifier was built in a prototype language. This work layed an important foundation for the subsequent work for the remaining time of the ERASMUS+ Internship in which the implementation was further extended and improved. Two exemplary use-cases were developed showing the potential of the layered approach.

One example of how the framework works can be seen in the accompanging code fragments. In \Cref{lst:code_after} we see a set of annotations that can be added to the base code of \Cref{lst:code_before}. We define three separate layers \texttt{state}, \texttt{types} and \texttt{liquid}. Internally, the framework will build a dependency graph thus allowing to verify properties independently from one another where appropriate.

\subsection{Deliverables and Publications}

The full implementation of the developed prototype will be published on GitHub\footnote{\url{https://github.com/LuAbelt/LayeredTypes}}. In addition a submission for a contributed talk was made to the TYPES2023 conference\footnote{\url{https://types2023.webs.upv.es/Index.html}} that is still pending a notification of acceptance or rejection. Additionally it is planned to potentially submit a paper to a relevant conference later in 2023.
